Robust variability of grid cell properties within individual grid modules enhances encoding of local space

Version 1

Abstract

Abstract

Although grid cells are one of the most well studied functional classes of neurons in the mammalian brain, the assumption that there is a single grid orientation and spacing per grid module has not been carefully tested. We investigate and analyze a recent large-scale recording of medial entorhinal cortex to characterize the presence and degree of heterogeneity of grid properties within individual modules. We find evidence for small, but robust, variability and hypothesize that this property of the grid code could enhance the ability of encoding local spatial information. Performing analysis on synthetic populations of grid cells, where we have complete control over the amount heterogeneity in grid properties, we demonstrate that variability, of a similar magnitude to the analyzed data, leads to significantly decreased decoding error, even when restricted to activity from a single module. Our results highlight how the heterogeneity of the neural response properties may benefit coding and opens new directions for theoretical and experimental analysis of grid cells.

The discovery of grid cells in medial entorhinal cortex (MEC) [1] has led to considerable experimental and computational work aimed at identifying their origin [2–12] and their function [13–21]. The organization of grid cells into discrete modules [22, 23], with particular grid properties (grid spacing and orientation) conserved within module, but not between modules, has fundamentally shaped this research. For instance, the increasing size and spacing of grid modules along the dorsal-ventral axis of MEC, by discontinuous jumps of a near constant ratio, has been argued to be optimal for encoding local spatial information [24, 25], when grid cell activity across all modules is integrated together [24–28]. This is despite the fact that the hippocampus, a downstream target of the MEC, receives inputs from only a portion of the dorsal-ventral axis [29]. The modularity of the grid system has also been proposed to simplify the wiring necessary for generating continuous attractor dynamics [2, 3, 5, 16], a computational mechanism theorized to underlie grid cells function that enjoys considerable experimental support [23, 30–34].

Much of this prior theoretical work has made the additional assumption that grid cell properties are identical, up to a phase shift, within a single module. However, the distributions of measured orientation and spacing show non-zero variability [22, 30, 34]. This could be due to finite recording time, neurophysiological noise, or the sensitivity of numerical methods used to fit grid properties. Alternatively, this variability could reflect underlying inhomogeneity, at a fine scale, within modules. Despite the fundamental way in which grid cell function has been informed by their modular organization, to the best of our knowledge, no characterization of the degree and robustness of variability in grid properties within individual modules has been performed.

If robust variability of grid properties does exist, then it is possible that this heterogeneity could be exploited to encode additional information about local space, reducing the requirement of integration across multiple grid modules (Fig. 1). In particular, when all grid cells have the same grid orientation and spacing, the joint population activity has a translational invariance that persists, even as larger populations of grid cells are considered (Fig. 1, “Fixed grid properties within module”). In contrast, if grid cells within a module have variability in their grid orientation and spacing, then – over a finite area – the translational invariance of the population activity is broken, and distinct patterns emerge in distinct locations of space (Fig. 1, “Variable grid properties within module”). As an example, the blue and green grid cells in Fig. 1 show the most overlap in the upper left half of the arena (denoted by cyan pixels), while the red and blue grid cells show the most overlap in the lower right portion of the arena (denoted by purple pixels). This is despite the fact that the variability in spacing and orientation is only on the order of a few centimeters and degrees, respectively.

Variability in grid cell properties within a module leads to enhanced encoding of local space.

When the activity of three idealized grid cells, all with the same grid spacing and orientation, are considered, the periodicity of the responses limits the amount of information conveyed about local space (Left column – “Fixed grid properties within module”). That is, there are multiple locations in physical space with identical population level activity. However, when three grid cells with variable grid spacing and orientation (in the realm of what is measured within individual grid modules – see Results), their joint activity contains considerably more information (Right column – “Variable grid properties within module”). This benefit of spatial inhomogeneity is expected to increase with larger populations of grid cells. Dashed squares in the joint activity map are enlarged below.

In this paper, we perform detailed analysis of recent state-of-the-art MEC electrophysiological recordings, which were made publicly available [34]. We characterize the variability of grid orientation and spacing within individual modules, and find evidence for small, but robust, variability. This variability is present whether a single value of grid orientation and spacing is assigned to each grid cell, or whether grid distortion [35, 36] is taken into account by considering three grid orientations and spacings independently. To assess the functional implications of this heterogeneity, we perform simulation experiments with synthetic, noisy grid cell populations, where we have complete control over the distribution of grid orientation and spacing. We find that the variability in grid cell orientation and spacing, at a similar degree as present in the data we analyze, leads to lower decoding error of local space when using the activity of a single module.

Taken together, our results challenge a long-held assumption and support a growing understanding of the spatial information encoded by grid cell populations [37–40]. Additionally, they encourage consideration of the broader benefits that multiple modules may provide, beyond the encoding of local space [41–43].

Results

Robust differences in grid cell properties within individual modules

To determine the extent to which variability of grid properties in individual modules exists, and to what extent this variability is a robust property of the grid code, we analyzed previously published MEC recordings [34], which include tens to hundreds of grid cells simultaneously recorded. This allows us to characterize the distribution of grid properties within a single grid module, to an extent not possible with other data sets.

For each grid cell, we compute the grid spacing (λ) and orientation (θ) by measuring properties associated with the six hexagonally distributed peaks of the spatial autocorrelogram (SAC), as traditionally performed [44] (Fig. 2A; see Materials and Methods). For clarity, we begin by focusing on a single module, recorded from an open field environment (recording identifier: Rat R, Day 1, Module 2 – R12). This module was picked for its long recording time (approximately 130 minutes of recorded activity, as compared to the other open field recordings that have 30 minutes or less of recorded activity) and for its large number of simultaneously recorded putative grid cells (N = 168). In this module, we find that grid cells with high grid scores (> 0.85; N = 74) have a range of grid orientation and spacing (example cells shown in Fig. 2B; distributions across module shown in Fig. 2E, F), with λ ranging from 65 cm to 90 cm and θ ranging from 2° to 9°. Overlaying the SAC of pairs of grid cells with similar grid spacing and different grid orientation (Fig. 2C) or vice versa (Fig. 2D) enables visualization of the extent of this variability.

Grid properties are variable within a single grid module (recording ID R12).

(A) Overview of the standard procedure used to calculate the grid spacing and orientation of a given grid cell. First, spike maps are computed by identifying the location of the animal at the time of each spike. Gray line denotes the trajectory of the rat, red dots denote locations of spikes. A rate map is constructed by binning space and normalizing by the amount of time the rat spent in each spatial bin. A spatial autocorrelogram (SAC) is computed and, after the center peak is masked out (white pixels in the center of the spatial autocorrelogram – leading to change in color scale), the grid properties are fit by measuring the length and angle of the three peaks closest to 0°. (B) Example grid cells from the same module (recording ID R12), with estimated grid score, orientation (θ), and spacing (λ). (C)–(D) SAC overlaid for two pairs of grid cells [from (B)]; one pair with different θ and similar λ (C) and the other with similar θ and different λ (D). (E)–(F) Distribution of θ (E) and λ (F) across all grid cells with grid score > 0.85 (N = 74).

Because individual grid fields can be cut-off by the boundaries of the environment, it is possible that computing the grid spacing from the SAC (which considers all grid fields) could lead to an under-estimate of λ, for some grid cells. To verify that the width of the distribution of grid spacing, within the same module, that we see is not due to the specifics of the SAC, we re-computed the grid spacing for all grid cells directly from their rate maps, considering only the three grid fields closest to the center of the environment (Fig. S2A; see Materials and Methods). We find that, while the grid spacing estimated from the SAC tends to be larger than the grid spacing estimated from the rate maps, a broad range of λ is again present (Fig. S2B, C).

To assess whether the heterogeneity of grid properties present in a single grid module is a robust feature or attributable to noise (either in the recording or the grid property fitting procedure), we measure the variability in grid orientation and spacing within a single grid cell and between pairs of grid cells. If the heterogeneity is explainable by noise, then we expect that the within-cell variability will be of the same size as the between-cell variability. In contrast, if the heterogeneity is a robust feature of the grid code, then we expect the within-cell variability will be significantly smaller than the between-cell variability.

To measure the within- and between-cell variability, we split the recording into evenly spaced 30 second bins, randomly assigning each temporal bin to one of two equal length halves and recomputing the grid properties for each half of the recording (Fig. 3A; see Materials and Methods). We set inclusion criteria to filter out cells that did not have consistent hexagonally structured SACs across splits of the recording (see Materials and Methods). While these requirements are strict (see Fig. S1, for percent of cells rejected), they set a conservative estimate on the amount of grid property variability, ensuring that we did not artificially inflate the variability due to weak grid cells. We do not find that the length of the temporal bin used to split the data has a large impact on the percent of cells accepted by this criteria (Fig. S3; see Materials and Methods).

Variability of grid properties is a robust feature of individual grid module (recording ID R12).

Schematic overview of approach used to compute the between- and within-cell variability of grid orientation and spacing. (B)–(C) Distribution of within- and between-cell variability of θ and λ, respectively. Note that the distribution is across all 100 random shuffles of the data into two halves. (D) Average within-cell variability of grid orientation , compared to average between-cell variability of grid spacing . (E) Same as (D), but for λ. 1 cell was excluded from (E) for visualization , but was included in non-parametric statistical analysis. For (D)–(E), N = 82.

We find that the distribution of within-cell variability of grid orientation and spacing is more concentrated around 0 than the between-cell variability, across all 100 random shuffles of the data into two halves. Comparing the average within- and between-cell variability of grid spacing and orientation reveals that nearly all of the grid cells that passed the criteria for inclusion (N = 82) exhibit more between-cell than within-cell variability (Fig. 3D, E): 95.1% grid cells for orientation ( denotes mean across cells) and 100% of grid cells for spacing .A Wilcoxon-Signed-Rank Test indicates that between-cell variability is significantly higher than within-cell variability for both orientation and spacing (Δθ: p < 0.001, Δλ : p < 0.001).

In keeping with convention, the reported grid cell properties are the average of those computed for each of the three independent axes in the SAC (Axis 1: aligned to ≈ 0°; Axis 2: aligned to ≈ 60°; Axis 3: aligned to ≈ −60° [45]). To ensure that this averaging is not contributing to the greater between-cell variability, we repeated the analysis above, restricting ourselves to each axis separately. The results again demonstrate that grid properties are significantly more robust within-cell than between-cell (Fig. 4A, B). For each axis, the average between-cell variability for every cell was significantly higher than the average within-cell variability (Fig. 4C, D), as reported by the Wilcoxon-Signed-Rank Test for orientation (p < 0.001, for all three axes) and for spacing (p < 0.001, for all three axes)

Variability of grid properties, restricted to the same axis, is a robust feature of individual grid module (recording ID R12).

Same analysis as in Fig. 3 (B) – (E), but for variability measured on each axis independently. For visualization, we exclude a small number of cells that were outside the axes limits, including 2, 5, and 10 cells for Axes 1–3, respectively (C); and 3, 4, and 3 cells for Axes 1–3, respectively (D); these cells were included in non-parametric statistical analyses.

Having demonstrated that grid cell properties are robustly heterogeneous in a single module, we proceed to analyze the remaining recordings in the data set (N = 420 cells across 8 modules; one module has no cells that pass our criteria) [34]. Although there are differences across recordings, we find larger between-than within-cell variability for grid orientation and spacing is present across all recordings (Fig. 5A, B; 80.0% of grid cells have greater between-than within-variability for orientation, and of grid cells have greater between-than within-variability for spacing, and ). A Wilcoxon-Signed-Rank Test finds that these differences are significant (Δθ: p < 0.001; Δλ : p < 0.001). We do not find evidence suggesting that the within-cell variability is influenced by grid score (Fig. 5C, D; linear regression for θ: R2 = 0.03, p = 0.55 Wald Test; linear regression for λ: R2 = 0.07, p = 0.14 Wald Test).

Within module grid property variability is a robust feature across modules.

(A) Average within-cell variability of grid orientation , compared to average between-cell variability of grid orientation for each cell (N = 420) across 8 modules (cells colored by their corresponding recording ID). The histogram above the plot shows the distribution of and the histogram to the right shows the distribution of . Same as (A), but for grid spacing. For visualization, 5 cells are excluded , but are included in non-parametric statistical analyses. Dashed gray lines show the population mean. (C)–(D) Average within cell variability of θ and λ (respectively), as a function of grid score. For visualization, 3 and 22 cells are excluded from (C)–(D), respectively, but are included in statistical analyses. Black solid line is linear regression, with R2 and p-value reported above.

Taken together, our analysis of large-scale MEC recordings demonstrates that grid cells in the same grid module do not have a single grid spacing and orientation, but instead have a restricted range of values around the module mean. This property is a robust feature that cannot be explained by noise from the recording or fitting procedures.

Variability in grid properties within single modules improves the encoding of local space

The variability of grid cell properties within individual grid modules, while statistically significant, is small in magnitude. Can a computational benefit in the encoding of local space be gained from this level of inhomogeneity? How sensitive might such a computational benefit be to the exact amount of variability present?

To address these questions, we generate populations of synthetic grid cells (see Materials and Methods), where we have complete control over the number of grid cells and their firing properties. For simplicity, we assume that all grid cells in our population have grid orientation and spacing sampled from Gaussian distributions, with means and and standard deviations σλ and σθ, respectively. Assigning each grid cell in our population a grid orientation and spacing, we are able to generate “ideal” rate maps [14]. Sampling from a Poisson process on these ideal rate maps, we generate noisy grid cell rate maps (Fig. 6A). Leveraging a simple linear decoder (see Materials and Methods), we can examine how decoding error of local space is affected by the number of grid cells in the population and the amount of variability (σλ and σθ).

Variability in grid properties enables improved decoding of local space from the activity of grid cells within a single module.

(A) Example noisy grid cell rate maps generated from a Poisson process. The size of the square arena is set to 1.5 m × 1.5 m to be consistent with what was used in the experimental set-up analyzed [34]. (B)–(C) Distribution of sampled grid spacing and orientation from synthetic population, when using , σλ = 5 cm, and σθ = 1°; compare to the distribution measured from real data (Fig. 2E, F). (D) Decoding error, as a function of grid cell population size, with populations having either no variability in grid properties (black line) or variability similar to what was present in the data analyzed (blue line). The solid line is the mean across 25 independent grid cell populations and the shaded area is ± standard deviation of the 25 independent populations. The dashed black line shows chance level decoding error. (E) Decoding error for synthetic populations and real data for up to N = 64 cells (red line). (F) Decoding error, over a grid of σθ and σλ values, for populations of N = 1024 grid cells. White star denotes values used in (D).

We begin by investigating the decoding capabilities of a synthetic module with properties similar to that of the experimentally recorded module that was analyzed in detail (recording identifier R12; Figs. 2, 3). We therefore set and . To determine an appropriate value for σλ and σθ, we subtract the mean between-cell variability by the mean within-cell variability ( and ) We interpret these values as the amount of variability in grid spacing and orientation that is not due to noise. Therefore, we set σθ = 1° and σλ = 5 cm. This amount of variability leads to similar sampled distributions of λ and θ as was found in the real data (compare Fig. 2E, F with Fig. 6B, C).

Applying the linear decoder to populations of synthetic grid cells, we find that decoding error decreases as the number of grid cells is increased (Fig. 6D, blue line). When N = 1024, the decoding error is ≈ 20 cm, substantially better than random decoding (Fig. 6D, dashed black line). As expected, this result is not seen in a synthetic population of grid cells with no variability in grid properties (σθ = 0° and σλ = 0 cm) (Fig. 6D, black line). In particular, while the restricted number of grid firing fields enables a decoding error smaller than chance, the population with fixed grid properties exhibits little change in decoding error with increasing numbers of grid cells. This demonstrates that the improved encoding is specific to populations with inhomogeneity in their grid spacing and orientation.

To validate that the synthetic population was a reasonable surrogate to the experimentally recorded data, we perform the same decoding analysis on grid cells from recording ID R12 (see Materials and Methods). As the real data is limited in the number of cells, we are only able to compare up to populations of N = 64 (Fig. 6E). However, in this restricted range, we find agreement between the synthetic population with grid property variability and the recorded data (Fig. 6E, compare red and blue lines). This strengthens our hypothesis that the observed amounts of inhomogeneity in grid spacing and orientation can lead to improved decoding of local space.

To determine the extent to which the decrease in decoding error depends the exact variability of grid spacing and orientation, we perform a grid search over 36 pairs of (σλ, σθ) values in [0°, 1°, …, 5°] × [0 cm, 1 cm, …, 5 cm]. As expected, when the variability of both grid properties is large, we find nearly 0 decoding error of local space (Fig. 6E, bottom right). However, there is additionally a range of σλ and σθ values that lead to improved decoding, including values smaller than used (Fig. 6E, above and to the left of the white star). We perform the same grid search on synthetic populations of grid cells with different spacings and (Fig. S4A, B), finding it again possible for the decoding error to drop well below that of the fixed population, if a sufficient amount of variability exists. For the synthetic module with smaller spacing , the existence of more grid fields in the 1.5 m × 1.5 m arena leads to greater complexity of interference patterns (Fig. 1), enabling the same amount of variability in grid spacing and orientation to lead to a sharper decrease in decoding error. Additionally, for the synthetic module with smaller spacing, a given value of σλ is a larger percentage of , as compared to the synthetic module with larger spacing.

Finally, we consider how decoding within a single module compares to decoding across two modules. When the two modules are consecutive (e.g., modules 2 and 3), their grid spacing, λ2 and λ3, has been experimentally found to be related via [22]. In such a case, decoding activity from populations with fixed grid properties leads to nearly 0 error (Fig. S5A, black line). The addition of variability in grid properties does not significantly change the behavior of the decoding (Fig. S5A, blue line). This suggests that small amounts of inhomogeneity may not disrupt the previously achieved theoretical bounds on decoding from multiple grid modules [24, 25, 28, 46]. However, if the two modules being decoded from are non-consecutive (e.g., modules 1 and 3) then the grid spacing can be related by an integer multiple, λ3 = 2λ1. In such a setting, the grid fields of the larger module are a subset of the grid fields of the smaller module, up to a rotation (due to the difference in orientation between the two modules), and we again find that variability in grid properties can improve decoding accuracy (Fig. S5B, compare black and blue lines). Taken together, these results suggest that individual grid modules can exhibit significant encoding of local space via heterogeneity in their grid properties, even when the extent of the variability in θ and λ is similar to that found in the analysis of the experimental recordings. This benefit can also improve encoding in cases when multiple, non-consecutive modules are considered.

Discussion

The multiple firing fields of grid cells, organized along a triangular lattice, has been historically interpreted as a limiting feature for encoding of local space. Particularly influential in shaping this view has been the discovery of the distribution of grid cells into distinct modules, with grid cell spacing (λ) and orientation (θ) preserved within, but not across, modules [22, 23], making integration across multiple modules necessary for spatial information to be decoded [24–28]. While evidence for discontinuity in the grid cell properties across modules is strong, the corollary assumption, that within-module values of λ and θ are identical (within the bounds of noise), has not been systematically studied.

Analyzing recently collected MEC recordings, we found the range of λ and θ values was large, with examples of grid cell pairs in the same module having over 7° difference in grid orientation and 20 cm difference in grid spacing (Fig. 2). Statistical analysis shows that the variability is more robust than expected from noise, for the majority of grid cells (Fig. 5). This was despite the fact that we used a very conservative criteria for assessing whether a grid cell was robust enough to include in the analysis.

Our comparison of within- and between-cell grid property variability was key to our argument, as it was for previous work using it to study the robustness of differences in peak grid field firing rates [39]. The absence of its use in the characterization of distribution of λ and θ [30] may be why the heterogeneity was not identified until now. We find that our analysis is robust to choices of parameters (Figs. S1, S3) and whether we treat each grid field independently or take the average across grid fields (Figs. 3, 4). This challenges the commonly held belief that the variability observed in the grid orientation and spacing is attributable solely to measurement noise.

We hypothesize that this variability may be used to increase the fidelity at which individual grid modules can encode local space. This idea is consistent with a large body of literature showing that heterogeneity in the responses of populations of neurons increases the robustness of encoding [47–50]. We indeed find, in noisy synthetic models of grid cells, that a level of variability in grid properties similar to what is quantified in the real data can be sufficient to accurately decode information of local space (Fig. 6). This benefit is increased with larger numbers of grid cells in the population (Fig. 6D), and is observed over a range of underlying variability values (Fig. 6F). We find that the improvement was most pronounced in modules with small grid spacing (Fig. S4A), although larger modules can see a decrease in decoding error for amounts of variability consistent with was was found in the analyzed data (Fig. S4B).

We note that our results are additionally aligned with recent findings of heterogeneity in maximum firing rates across individual grid fields [37–39] and grid sheering [35, 36, 40]. Our work further demonstrates that, even in the absence of these perturbations, individual grid modules may encode considerably more local spatial information than previously believed.

Finally, models of the formation of orientation maps in visual cortex have demonstrated that slight angular offsets of retinal mosaics, along which retinal receptive fields are organized, can generate complex patterns [51], similar to those found in visual cortex orientation maps [52]. Our results indicate that grid cells in MEC may be capable of leveraging a similar computational principle, suggesting that mosaic patterns might be a broadly utilized feature in neural coding [53].

Limitations

While the data set we analyzed [34] represents an advance in the ability to simultaneously record from tens to hundreds of putative grid cells, across grid modules, the MEC remains a challenging brain region to access for large-scale neurophysiological experiments. Indeed, with our conservative inclusion criteria, we were ultimately limited by having only 420 grid cells included in our analysis. Future work can perform more detailed and complete characterizations of grid property heterogeneity, as new neurotechnologies that enable larger yield of grid cells are developed [54, 55].

Our decoding analysis, while demonstrating the possibility that variability in grid properties can be leveraged by single grid modules to enhance the encoding of local spatial information, made several simplifying assumptions: Poisson sampling for spike rates; linear decoding; normal distribution of grid properties. While comparison to real data showed that these assumptions are reasonable (Fig. 6E), future work can assess the extent to which these restrictions can be lifted, while still enabling individual grid modules to have low decoding error.

Open questions

The heterogeneity in grid properties we characterize motivates the investigation of several new lines of research. Because these are directions that we believe to be fruitful for the field as a whole, we outline them below, with our hypotheses for possible answers.

Q1: What causes grid property variability?

A natural first direction to address is identifying the source of the heterogeneity in grid cell properties we observe. One hypothesis is that this could be driven by “defects” in the specific connectivity pattern that is needed for a continuous attractor. However, whether stable path integration could occur, in the face of such defects, is unclear. As an alternative, the coupling between hippocampus and MEC, which has been shown to lead to variability in grid field firing rates (as experimentally observed) [39, 56], may lead to differences in grid orientation and spacing. In particular, a previous computational model that learned grid cells from place cell input, using non-negative principal component analysis, has shown that place field width affects grid cell orientation and spacing [6]. Heterogeneity in the spatial coding properties of place cells that has been found along the transverse axis of CA3 [57–59], suggesting there may be a systematic differences in the place field widths of hippocampal inputs to MEC grid cells. Further, it was shown that this place-to-grid cell computational model has a linear relationship between place field width and grid spacing, and a non-monotonic relationship between place field width and grid orientation [6]. These may explain why we find stronger average variability in grid spacing than grid orientation (Fig. 5A, B).

Q2: How does grid property variability affect continuous attractor network structure?

Continuous attractor network models of grid cells [2, 3, 5, 16] enjoy considerable experimental support [23, 30–34], making them one of the “canonical” models in neuroscience. However, these models make use of the assumption that all the grid cells in a given module have the same grid orientation and spacing to simplify the network connectivity. In particular, by assuming equal grid orientation and spacing, it becomes possible to arrange grid cells in a two-dimensional space spanned by their phases (i.e., a “neural sheet” [16]). Neurons close in this space are wired with excitatory connections and neurons far in this space are wired with inhibitory connections. As the data set we analyzed was found to provide strong support for the basic predictions of continuous attractor networks (i.e., toroidal topology of the activity manifold) [34], we do not view our results as directly challenging these models. However, considering how these models can accommodate the observed variability is an important future direction. It is possible that these modifications are learnable, via some plasticity mechanism. These solutions may lead to geometric effects on the underlying manifold, which could be explored by using computational tools [60], as well as affect the dynamical properties of the neural population activity, which could be quantified by data-driven dynamical systems methods [61–63]. We hypothesize that the degree in variability of grid spacing and orientation may strike a balance between being small enough to keep the continuous attractor network structure stable, but large enough to enable encoding of local information of space.

Q3: How does grid property variability shape hippocampal representations?

The projections from MEC to hippocampus suggest that the variability in grid properties may influence hippocampal representations (even if grid cells do not comprise the majority of its inputs). We consider two possible ways in which this may happen. First, given that grid cells have been reported to maintain their grid spacing and orientation across exposures to new environments, while undergoing a change in their grid phase [30, 64], the integration across multiple modules has been necessary to explain place field remapping. However, grid phase plays an important role in generating the specific complex interference patterns that emerge when considering the joint activity of grid cells with variable grid properties (Fig. 1). Thus the reported changes in phase may be sufficient to generate large (and seemingly random) changes in local spatial information conveyed by grid cells to hippocampal cells. This could drive additional changes in the local spatial information projected to hippocampus, as well as explain the significant differences in correlation structure between CA1 neurons across different environments [65].

And second, recent work on hippocampal place field drift [66–70] has demonstrated that there is a significant change in the place field location across time, especially in CA1. One possible source of this phenomenon is the reported instability of dendritic spines on the apical dendrites of CA1 place cells [59, 71– 73], ostensibly leading to changes in the MEC inputs to these neurons. However, if grid cells across multiple modules are necessary for local spatial information, turnover in synaptic input is unlikely to cause large changes in the spatial preferences of CA1 neurons, as integration over several scales should provide stable encoding properties. In contrast, different subpopulations of grid cells with variable grid properties can lead to differences in the local spatial information encoded by their joint activity, even if they come from a single module, possibly influencing the spatial preferences of CA1 place cells.

Q4: Why are there multiple modules?

Given that our results demonstrate the ability of single grid modules to encode information about local space – a feat previously believed to be possible only if activity from multiple grid modules was integrated together – why is does MEC have multiple modules? While the variability removes the necessity for encoding local space with multiple modules, higher fidelity representations is achievable by integrating across multiple modules (Fig. S5). In addition, the use of grid cells beyond spatial navigation [74–79], where hierarchical representations are important [42, 43], may be a sufficient implicit bias for the formation of multiple modules [80]. In particular, encoding information at multiple distinct scales is critical for multi-scale reasoning, a cognitive function grid cells may support.

Materials and Methods

Electrophysiology recordings

The neural activity analyzed in this paper comes from a publicly available data set, which has previously been described in detail [34]. We provide brief summary of the methodology and the experimental paradigms used during the recordings.

Three male rats (Long Evans – Rats Q, R, and S) were implanted with Neuropixels silicon probes [81, 82]. These probes were targeted at the MEC-parasubiculum region and the surgery was performed as described previously [32, 82]. After three hours of recovery, recordings were performed.

In the recordings analyzed, the rats foraged for randomly dispersed corn puffs in a 1.5 × 1.5 m2 square open field arena, with walls of height 50 cm. The rats were familiar with the environment and task, having trained 10–20 times prior to the implantation. The rats were food restricted to motivate their foraging, being kept at a minimum of 90% of their original body weight (300–500 grams).

All procedures in the original study were approved by the Norwegian Food and Safety Authority and done in accordance with the Norwegian Animal Welfare Act and the European Convention for the Protection of Vertebrate Animals used for Experimental and Other Scientific Purposes.

Electrophysiology post-processing

The neural activity analyzed in this paper was post-processed, before made publicly available. We describe, in brief, the post-processing performed [34], as well as the post-processing we performed on the downloaded data.

Spike sorting, via KiloSort 2.5 [82], was applied to the data recorded from the Neuropixel probes. Individual units were deemed putative cells if their average spike rate was in the range of 0.5–10Hz, and 99% of their interspike intervals were greater than 2 ms.

For each putative cell, rate maps were constructed by averaging the activity at binned spatial positions in the open field arena. This raw rate map was smoothed, using a Gaussian kernel. The autocorrelation of these rate maps were computed, and a grid score calculated, as described previously [44]. To determine which putative cells were grid cells, and to which module they belonged, autocorrelograms of raw rate maps were constructed, with some of the bins corresponding to the closest and furthest distances removed. These “cropped” rate maps were then clustered by using UMAP [83] to project the data onto a two-dimensional subspace. All the cells in the non-largest cluster were found to have similar grid spacing and orientation, as well as high grid scores. They were therefore deemed putative grid cells. All cells in these clusters had their spike trains placed in the public repository from which we downloaded the data.

From the downloaded spike trains, we constructed rate maps and autocorrelograms in a similar manner, using code made publicly available [8] (Fig. 2).

Computing grid score, orientation and spacing from the spatial autocorrelogram

Grid spacing and grid orientation were computed according to standard methods described in detail previously [22, 84]. Briefly, the goal of the procedure is to identify the location of the six nearest fields in the spatial autocorrelogram (SAC) (Fig. 2A). This was achieved by performing the following steps. First, the SAC was smoothed using a 2D Gaussian filter with σ = 1. Second, the center of the SAC was excluded by applying a mask. Because the size of the SAC’s central peak changes for every module, the radius of such mask was re-computed for each module. Third, we thresholded the SAC by applying an extended-maxima transform ndimage.maximum_filter. Fourth, we identified the center of every field by using the function scipy.stats.find_peaks.

Once the peaks of every field had been found, we computed the location of every peak in polar coordinates. We then selected the 6 peaks that were closest to the center of the SAC, based on the computed radial components. Because every SAC is symmetric, we considered for further analysis the 3 peaks closest to the X axis in angular distance (Axis 1, 2, and 3 [45]). Grid spacing was computed as the arithmetic mean of the radial component of the 3 peaks (except when each peak was analyzed separately – Fig. 4). Given that the SAC dimensions are twice of that of the real arena, we multiplied the SAC radial mean by a factor of 2. Grid orientation was computed as the angular mean of orientations (relative to the x-axis) of the three peaks (except when each peak was analyzed separately – Fig. 4).

In order to ensure that subsequent analysis was performed only on cells whose SAC’s could be well described by hexagonal structure, we imposed the following constraints: 1) the relative angle between two peaks could not be < 30°; 2) the relative angle between two peaks could not be > 90°; 3) the values for grid spacing between the peaks could not be substantially different (the ratio of spacings between any two peaks must be > 0.5 and < 2). Cells that did not meet these criteria were determined to have “Poor grid fit” and were rejected from all subsequent analysis. The percentage of all cells that were removed by this inclusion criteria is shown in Fig. S1. We note that cells from one of the nine modules in the publicly available data set that we analyzed had 98.4% of cells rejected by this criteria (recording identifier – R23). We therefore did not include it in any subsequent analysis.

To compute the grid score of recorded MEC cells, we made use of previously published code [8], that is based on metrics that have become standards in quantifying grid cell properties[44].

For analysis of the distribution of grid properties (Fig. 2E, F), we included only grid cells with grid scores greater than 0.85. This was done to demonstrate that variability was present even in cells that exhibit robust grid cell properties. In the subsequent analyses, an alternative criteria is used, which considers the reliability of the grid responses (see below).

Computing spacing from the rate maps

In order to characterize the spacing with a method that is less susceptible to the effects of having grid fields at the boundaries of the environment (as the SAC method could be), we compute the spacing using the rate maps of individual grid cells (Fig S2A). Once the rate map was computed, we smoothed it with a Gaussian filter, setting σ = 1.5. Then, using the function scikit-image.feature.peak_local_max we extract the position of the center of each firing field. Because our goal with this analysis is to show that the fields in the border do not impact our analysis with the SAC, we restrict ourselves to the 3 peaks that are closest to the center of the arena (Fig S2A). We compute the distances between those three peaks to each other and report the spacing as the average of the distances measured.

Within and between cells splits

To characterize the within- and between-cell variability of grid spacing and orientation, we employed the following approach. First, we split the data into bins of fixed length (30 seconds). From this, we randomly assigned each interval to one of two blocks (denoted as blocks A and B), with exactly half the total number of intervals in each block. For each grid cell, we computed the grid spacing [and] and orientation [ and], from the data in each block. The within-cell differences of grid spacing and orientation was determined as and . Each grid cell’s properties were also compared those of another grid cell, with the match being made using random sampling without replacement. These comparisons were determined as and . This process was repeated 100 times (referring to each iteration as a “shuffle”), per recording. Examples of splits of the data, for different shuffles, is schematically illustrated in Fig. 3A). The resulting distributions of Δθwithin, Δλwithin, Δθbetween, and Δλbetween, across grid cells and splits of the data were then compared. If the within-cell variability in grid spacing and orientation was smaller than between-cell variability, we concluded that the variability of grid cell properties was a robust feature of the data and not due to noise.

When performing this shuffle analysis, we found that some cells, despite having a good grid fit when all the data was considered, did not have SACs that were well described by hexagonal structure when the data was split in half. We viewed this a manifestation of unreliable grid coding and a possible confound in our quantification of variability. As such, we introduced a new inclusion criteria (replacing that of requiring a grid score of > 0.85), only considering cells that had poor grid fits on < 5% of all shuffles. The percentage of all cells that were removed by this inclusion criteria is shown in Fig. S1. In general, we found that cells with high grid score were reliable, although there were exceptions. Additionally, we found that size of the bin used for splitting the data did not significantly affect the percent of cells with good grid fits (passed the prior inclusion criteria) that were considered reliable (Fig. S3).

Synthetic grid cells

To study how variability in grid cell properties might endow the grid code with computational advantages, we generated synthetic grid cell rate maps, so that we could have complete control over the distribution of their properties. These synthetic grid cell rate maps were constructed as follows.

First, the lengths of each dimension the “arena” within which the simulated grid cells exist (Lx and Ly) were set. Then, for N ∈ ℕ grid cells, the grid spacing and orientation were sampled via λ(i) ∼ 𝒩 (µλ, σλ) and θ(i) ∼ 𝒩 (µθ, σθ) where 𝒩 (µ, σ) is a normal distribution with mean µ and variance σ2. Grid phase was sampled as ϕ(i) 𝒰 ([0, Lx] × [0, Ly]), where ϕ is a two dimensional vector, with first component uniformly sampled from [0, Lx] and second component uniformly sampled from [0, Ly]. To construct a population with no variability in grid properties (to use as a control), we set σλ = 0 m. and σθ = 0°.

For each grid cell, we generated idealized grid responses by summing three two-dimensional sinusoids [14], such that the activity at x = (x, y) ∈ [−Lx/2, Lx/2] ×[−Ly/2, Ly/2] is given by where is the maximal firing rate, R(θi) is the two-dimensional rotation matrix, with rotation θi, and kj are the wave vectors with 0°, 60° and 120° angular differences [i.e.,. To match the recorded neural data, where individual grid cells have distinct maximal firing rates,. We enforced to be within [2, 30], by setting any sampled values outside of this range to the boundary values (i.e., 2 or 30).

To determine the extent to which local spatial information can be decoded from the activity of populations of grid cells with different degrees of variability in their grid properties, we performed the following analysis.

We generated noisy synthetic spike rates of N grid cells by assuming a Poisson process and sampling using the idealized rate maps (Eq. 1). More concretely, the activity of grid cell i at position (x, y) was assumed to be a random variable with a Poisson distribution, whose mean was X(i)(x, y). Thus, the probability of observing spikes, at position (x, y), is given by

Linear decoding synthetic data

For a given resolution of the arena, we generated , where j = 1, …, 10. That is, we constructed 10 noisy rate maps. We performed cross-validated decoding by averaging across 9 of the 10 rate maps, to get an average rate map . For sake of simplicity, consider . To decode local position from the held out noisy rate map [e.g.], we multiplied the activity at each position by all the positions in the average rate map, taking the sum and assigning the decoded position as that with the largest value. The Euclidean distance between the decoded position and the true position is considered the error. This was performed 10 times (holding out each rate map once) and the average error across all positions in the environment was then averaged across all 10 of the validation splits.

Linear decoding experimental data

To decode the electrophysiological data [34], we sampled subpopulations of grid cells from the N = 82 units (recording ID R12) that passed the criteria described previously. For each subpopulation, we split the recorded activity into 10 evenly spaced temporal bins, and constructed average ratemaps for each bin. This is consistent with what was done in decoding the synthetic data. All ratemaps, for each cell, were normalized by the maximal activity. We then randomly chose 5 of the 10 time bins to serve as the “train” data, the remaining 5 served as the “test” data. The rate maps were averaged and then the linear decoder was applied. We sampled 10 different choices of splitting the data and 25 choices of subpopulation.

eLife Assessment

This valuable study characterizes the variability in spacing and direction of entorhinal grid cells and shows how this variability can be used to disambiguate locations within an environment. These claims are supported by solid evidence, yet some aspects of the methodology should be clarified. This study will be of interest to neuroscientists working on spatial navigation and, more generally, on neural coding.

Reviewer #1 (Public review):

Summary:

The present paper by Redman et al. investigated the variability of grid cell properties in the MEC by analyzing publicly available large-scale neural recording data. Although previous studies have proposed that grid spacing and orientation are homogeneous within the same grid module, the authors found a small but robust variability in grid spacing and orientation across grid cells in the same module. The authors also showed, through model simulations, that such variability is useful for decoding spatial position.

Strengths:

The results of this study provide novel and intriguing insights into how grid cells compose the cognitive map in the axis of the entorhinal cortex and hippocampus. This study analyzes large data sets in an appropriate manner and the results are solid.

Weaknesses:

A weakness of this paper is that the scope of the study may be somewhat narrow, as this study focused only on the variability of spacing and orientation across grid cells. I would suggest some additional analysis or discussion that might increase the value of the paper.

(1) Is the variability in grid spacing and orientation that the authors found intrinsically organized or is it shaped by experience? Previous research has shown that grid representations can be modified through experience (e.g., Boccara et al., Science 2019). To understand the dynamics of the network, it would be important to investigate whether robust variability exists from the beginning of the task period (recording period) or whether variability emerges in an experience-dependent manner within a session.

(2) It is important to consider the optimal variability size. The larger the variability, the better it is for decoding. On the other hand, as the authors state in the Discussion, it is assumed that variability does not exist in the continuous attractor model. Although this study describes that it does not address how such variability fits the attractor theory, it would be better if more detailed ideas and suggestions were provided as to what direction the study could take to clarify the optimal size of variability.

Reviewer #2 (Public review):

Summary:

This paper presents an interesting and useful analysis of grid cell heterogeneity, showing that the experimentally observed heterogeneity of spacing and orientation within a grid cell module can allow more accurate decoding of location from a single module.

Strengths:

I found the statistical analysis of the grid cell variability to be very systematic and convincing. I also found the evidence for enhanced decoding of location based on between-cell variability within a module to be convincing and important, supporting their conclusions.

Weaknesses:

(1) Even though theoreticians might have gotten the mistaken impression that grid cells are highly regular, this might be due to an overemphasis on regularity in a subset of papers. Most experimentalists working with grid cells know that many if not most grid cells show high variability of firing fields within a single neuron, though this analysis focuses on between neurons. In response to this comment, the reviewers should tone down and modify their statements about what are the current assumptions of the field (and if possible provide a short supplemental section with direct quotes from various papers that have made these assumptions).

(2) The authors state that "no characterization of the degree and robustness of variability in grid properties within individual modules has been performed." It is always dangerous to speak in absolute terms about what has been done in scientific studies. It is true that few studies have had the number of grid cells necessary to make comparisons within and between modules, but many studies have clearly shown the distribution of spacing in neuronal data (e.g. Hafting et al., 2005; Barry et al., 2007; Stensola et al., 2012; Hardcastle et al., 2015) so the variability has been visible in the data presentations. Also, most researchers in the field are well aware that highly consistent grid cells are much rarer than messy grid cells that have unevenly spaced firing fields. This doesn't hurt the importance of the paper, but they need to tone down their statements about the lack of previous awareness of variability (specific locations are noted in the specific comments).

(3) The methods section needs to have a separate subheading entitled: How grid cells were assigned to modules" that clearly describes how the grid cells were assigned to a module (i.e. was this done by Gardner et al., or done as part of this paper's post-processing?

Reviewer #3 (Public review):

Summary:

Redman and colleagues analyze grid cell data obtained from public databases. They show that there is significant variability in spacing and orientation within a module. They show that the difference in spacing and orientation for a pair of cells is larger than the one obtained for two independent maps of the same cell. They speculate that this variability could be useful to disambiguate the rat position if only information from a single module is used by a decoder.

Strengths:

The strengths of this work lie in its conciseness, clarity, and the potential significance of its findings for the grid cell community, which has largely overlooked this issue for the past two decades. Their hypothesis is well stated and the analyses are solid.

Weaknesses:

On the side of weaknesses, we identified two aspects of concern. First, alternative explanations for the main result exist that should be explored and ruled out. Second, the authors' speculation about the benefits of variability in angle and spacing for spatial coding is not particularly convincing, although this issue does not diminish the importance or impact of the results.

Major comments:

(1) One possible explanation of the dispersion in lambda (not in theta) could be variability in the typical width of the field. For a fixed spacing, wider fields might push the six fields around the center of the autocorrelogram toward the outside, depending on the details of how exactly the position of these fields is calculated. We recommend authors show that lambda does not correlate with field width, or at least that the variability explained by field width is smaller than the overall lambda variability.

(2) An alternative explanation could be related to what happens at the borders. The authors tackle this issue in Figure S2 but introduce a different way of measuring lambda based on three fields, which in our view is not optimal. We recommend showing that the dispersions in lambda and theta remain invariant as one removes the border-most part of the maps but estimating lambda through the autocorrelogram of the remaining part of the map. Of course, there is a limit to how much can be removed before measures of lambda and theta become very noisy.

(3) A third possibility is slightly more tricky. Some works (for example Kropff et al, 2015) have shown that fields anticipate the rat position, so every time the rat traverses them they appear slightly displaced opposite to the direction of movement. The amount of displacement depends on the velocity. Maps that we construct out of a whole session should be deformed in a perfectly symmetric way if rats traverse fields in all directions and speeds. However, if the cell is conjunctive, we would expect a deformation mainly along the cell's preferred head direction. Since conjunctive cells have all possible preferred directions, and many grid cells are not conjunctive at all, this phenomenon could create variability in theta and lambda that is not a legitimate one but rather associated with the way we pool data to construct maps. To rule away this possibility, we recommend the authors study the variability in theta and lambda of conjunctive vs non-conjunctive grid cells. If the authors suspect that this phenomenon could explain part of their results, they should also take into account the findings of Gerlei and colleagues (2020) from the Nolan lab, that add complexity to this issue.

(4) The results in Figure 6 are correct, but we are not convinced by the argument. The fact that grid cells fire in the same way in different parts of the environment and in different environments is what gives them their appeal as a platform for path integration since displacement can be calculated independently of the location of the animal. Losing this universal platform is, in our view, too much of a price to pay when the only gain is the possibility of decoding position from a single module (or non-adjacent modules) which, as the authors discuss, is probably never the case. Besides, similar disambiguation of positions within the environment would come for free by adding to the decoding algorithm spatial cells (non-hexagonal but spatially stable), which are ubiquitous across the entorhinal cortex. Thus, it seems to us that - at least along this line of argumentation - with variability the network is losing a lot but not gaining much.

(5) In Figure 4 one axis has markedly lower variability. Is this always the same axis? Can the authors comment more on this finding?

(6) The paper would gain in depth if maps coming out of different computational models could be analyzed in the same way.

(7) Similarly, it would be very interesting to expand the study with some other data to understand if between-cell delta_theta and delta_lambda are invariant across environments. In a related matter, is there a correlation between delta_theta (delta_lambda) for the first vs for the second half of the session? We expect there should be a significant correlation, it would be nice to show it.

Author response:

The following is the authors’ response to the original reviews.

We thank the reviewers for their time and thoughtful comments. We believe that the further analyses suggested have made the results clearer and more robust. Below, we briefly highlight the key points addressed in the revision and the new evidence supporting them. Then, we address each reviewer’s critiques point-by-point.

- Changes in variability with respect to time/experience

Both reviewers #1 and #3 asked whether the variability in grid properties observed was dependent on time or experience. This is an important point, given that such a dependence on time could lead to interesting hypotheses about the underlying dynamics of the grid code. However, in the new analyses we performed, we do not observe changes in grid variability within a session (Fig S5 of the revised manuscript), suggesting that the grid variability seen is constant within the timescale of the data set.

- The assumption of constant grid parameters in the literature

Reviewer #2 pointed out that it had been appreciated by experimentalists that grid properties are variable within a module. We agree that we may have overstated the universality of this assumption in the original manuscript, and we have toned down the language in the revision. However, we note that many previous theoretical studies assumed these properties to be constant, within a given module. We provide some examples below, and have added evidence of this assertion, with citations to the theoretical literature, to the revised manuscript .

- Additional sources of variability

Reviewer #3 pointed out additional sources that might explain the variability observed in the paper (beyond time and experience). These sources include: field width, border location, and the impact of conjunctive cells. We have run additional analyses and have found no significant impact on the observed variability from any of these factors. We believe that these are important controls, and have added them to the manuscript (Fig S4-S7 of the revised manuscript)

- Analysis of computational models

Reviewer #3 noted that our results could be strengthened by performing similar analyses on the output of computational models of grid cells. This is a good idea. We have now measured the variability of grid properties in a recent normative recurrent neural network (RNN) model that develops grid cells when trained to perform path integration (Sorscher et al., 2019). This model has been shown to develop signatures of a 2D toroidal attractor (Sorscher et al., 2023) and achieves a high accuracy on a simple path integration task. Interestingly, the units with the greatest grid scores also exhibit a range of grid spacings and grid orientations (Fig S8 of the revised manuscript). Furthermore, by decreasing the amount of sparsity (through decreasing the weight decay regularization), we found an increase in the variability of the grid properties. This analysis demonstrates a heretofore unknown similarity between the RNN models trained to perform path integration and recorded grid cells from MEC. It additionally provides a framework for computational analysis of the emergence of grid property variability.

Reviewer #1:

(1) Is the variability in grid spacing and orientation that the authors found intrinsically organized or is it shaped by experience? Previous research has shown that grid representations can be modified through experience (e.g., Boccara et al., Science 2019). To understand the dynamics of the network, it would be important to investigate whether robust variability exists from the beginning of the task period (recording period) or whether variability emerges in an experience-dependent manner within a session.

This is an interesting question that was not addressed in the paper. To test this, we performed additional analysis to resolve whether the variability changes across a session.

Using a sliding window, we have measured changes in variability with respect to recording time (Fig S5A). To this end, we compute grid orientation and spacing over a time-window whose length is half the total length of the recording. From the population distribution of orientation and spacing values, we compute the standard deviation as a measure of variability. We repeat the same procedure, sliding the window forward until the variability for the second half of the recording is computed.

We applied this approach to recording ID R12 (the same as in Figs 2-4) given that this recording session was significantly longer than the rest (nearly two hours). Results are shown in Fig S5B-C. For both orientation and spacing, no changes of variability with respect to time can be observed. Similar results were found for other modules (see caption of Fig S5 for statistics).

We also note that the rats were already familiarized with the environment for 10-20 sessions prior to the recordings, so there may not be further learning during the period of the grid cell recordings. No changes in variability can be seen in Rat R across days (e.g., in Fig 5B R12 and R22 have similar distributions of variability). However, we note that it may be possible that there are changes in grid properties at time-scales greater than the recordings.

(2) It is important to consider the optimal variability size. The larger the variability, the better it is for decoding. On the other hand, as the authors state in the

Discussion, it is assumed that variability does not exist in the continuous attractor model. Although this study describes that it does not address how such variability fits the attractor theory, it would be better if more detailed ideas and suggestions were provided as to what direction the study could take to clarify the optimal size of variability.

We appreciate this suggestion and agree that more discussion is warranted on how our results can be reconciled with previously observed attractor dynamics. To explore this, we studied the recurrent neural network (RNN) model from Sorscher et al. (2019), which develops grid responses when trained on path integration. This network has previously been found to develop signatures of toroidal topology (Sorscher et al., 2023), yet we find its grid responses also contain heterogeneity in grid properties (Fig S8). By decreasing the strength of the weight decay regularization (which leads to denser connectivity in the recurrent layer), we find an increase in the grid property variability. Interestingly, decreasing the weight decay regularization has been previously found to lead to weaker grid responses and worse ability of the RNN to perform path integration on environments larger than it was trained on. This approach not only provides preliminary evidence to our claim that too much variability can lead to weaker continuous attractor structure, but also provides a modeling framework with which future work can explore this question in more detail. We have added discussion of this issue to the manuscript text (Discussion).

Reviewer #2:

(1) Even though theoreticians might have gotten the mistaken impression that grid cells are highly regular, this might be due to an overemphasis on regularity in a subset of papers. Most experimentalists working with grid cells know that many if not most grid cells show high variability of firing fields within a single neuron, though this analysis focuses on between neurons. In response to this comment, the reviewers should tone down and modify their statements about what are the current assumptions of the field (and if possible provide a short supplemental section with direct quotes from various papers that have made these assumptions).

We agree that some experimentalists are aware of variability in the recorded grid response patterns and that this work may not come as a complete surprise to them. We have toned down our language in the Introduction, changing “our results challenge a long-held assumption” to “our results challenge a frequently made assumption in the theoretical literature”. Additionally, we have added a caveat that “experimentalists have been aware” of the observed variability in grid properties.

We would like to emphasize that the lack of work carefully examining the robustness of this variability has prevented a firm understanding of whether this is an inherent property of grid cells or due to measurement noise. The impact of this can be seen in theoretical neuroscience work where a considerable number of articles (including recent publications) start with the assumption that all grid cells within a module have identical properties, with the exception of phase shift and noise. We have now cited a number of these papers in the Introduction, to provide specific references. To further illustrate the pervasiveness of this assumption being explicitly made in theoretical neuroscience, below we provide quotes from a few important papers:

“Cells with a common spatial period also share a common grid orientation; their responses differ only by spatial translations, or different preferred firing phases, with respect to their common response period” (Sreenivasan and Fiete, 2011)”

“Grid cells are organized into discrete modules; within each module, the spatial scale and orientation of the grid lattice are the same, but the lattice for different cells is shifted in space.” (Stemmler et al., 2015)”

“Recently, it was shown that grid cells are organized in discrete modules within which cells share the same orientation and periodicity but vary randomly in phase” (Wei et al., 2015)”

“...cells within one module have receptive fields that are translated versions of one another, and different modules have firing lattices of different scales and orientations” (Dorrell et al., 2023)”

In these works, this assumption is used to derive properties relating to the computational properties of grid cells (e.g., error correction, optimal scaling between grid spacings in different modules).

In addition, since grid cells are assumed to be identical in the computational neuroscience community, there has been little work on quantifying how much variability a given model produces. This makes it challenging to understand how consistent different models are with our observations. This is illustrated in our analysis of a recent recurrent neural network (RNN) model of grid cells (Fig S8), which does exhibit variability.

(2) The authors state that "no characterization of the degree and robustness of variability in grid properties within individual modules has been performed." It is always dangerous to speak in absolute terms about what has been done in scientific studies. It is true that few studies have had the number of grid cells necessary to make comparisons within and between modules, but many studies have clearly shown the distribution of spacing in neuronal data (e.g. Hafting et al., 2005; Barry et al., 2007; Stensola et al., 2012; Hardcastle et al., 2015) so the variability has been visible in the data presentations. Also, most researchers in the field are well aware that highly consistent grid cells are much rarer than messy grid cells that have unevenly spaced firing fields. This doesn't hurt the importance of the paper, but they need to tone down their statements about the lack of previous awareness of variability (specific locations are noted in the specific comments).

We have toned down our language in the Introduction. However, we note that our point that no detailed analysis had been done on measuring the robustness of this variability stands. Thus, for the general community, it has not been clear whether this previously observed variability is noise or a real feature of the grid code.

(3) The methods section needs to have a separate subheading entitled: How grid cells were assigned to modules" that clearly describes how the grid cells were assigned to a module (i.e. was this done by Gardner et al., or done as part of this paper's post-processing?

We thank the reviewer for pointing out this missing information. We have added a new subsection in the Materials and Methods section, entitled “Grid module classification” to clarify how the grid cells are assigned to modules. In short, this was done by Gardner et al. (2022) using an unsupervised clustering approach that was viewed as enabling a less biased identification of modules. We did not perform any additional processing steps on module identity.

Reviewer #3:

(1) One possible explanation of the dispersion in lambda (not in theta) could be variability in the typical width of the field. For a fixed spacing, wider fields might push the six fields around the center of the autocorrelogram toward the outside, depending on the details of how exactly the position of these fields is calculated. We recommend authors show that lambda does not correlate with field width, or at least that the variability explained by field width is smaller than the overall lambda variability.

We agree that this option had not been carefully ruled out by our previous analyses. To tackle this question, we compute the field width of a given cell using the value at the minima of its spatial autocorrelogram (Fig S4A-B). For all cells in recording ID R12, there is a non-significant negative linear correlation between grid field width and between-cell variability (Fig S4C) . The variability explained by the width of the field is 4% of the variability, as indicated by the R2 value of the linear fit. Similar results were found for all other modules (see caption of Fig S4C for statistics). Therefore, we do not think that grid field width explains spacing variability.

(2) An alternative explanation could be related to what happens at the borders. The authors tackle this issue in Figure S2 but introduce a different way of measuring lambda based on three fields, which in our view is not optimal. We recommend showing that the dispersions in lambda and theta remain invariant as one removes the border-most part of the maps but estimating lambda through the autocorrelogram of the remaining part of the map. Of course, there is a limit to how much can be removed before measures of lambda and theta become very noisy.

We have performed additional analysis to explore the role of borders in grid property variability. To do so, we have followed the suggestion by the reviewer and have re-analyzed grid properties from the autocorrelogram when the border-most part of the maps are removed (Fig S6A-B). For all modules, we do not see any changes in variability (computed as the standard deviation of the population distribution) for either orientation or spacing. As predicted by the reviewer, after removing about 25% of the border-most part of the environment we start seeing changes in variability, as measures of theta and lambda become noisy and computed over a smaller spatial range. This result holds for all other modules (Fig S6C-D).

(3) A third possibility is slightly more tricky. Some works (for example Kropff et al, 2015) have shown that fields anticipate the rat position, so every time the rat traverses them they appear slightly displaced opposite to the direction of movement. The amount of displacement depends on the velocity. Maps that we construct out of a whole session should be deformed in a perfectly symmetric way if rats traverse fields in all directions and speeds. However, if the cell is conjunctive, we would expect a deformation mainly along the cell's preferred head direction. Since conjunctive cells have all possible preferred directions, and many grid cells are not conjunctive at all, this phenomenon could create variability in theta and lambda that is not a legitimate one but rather associated with the way we pool data to construct maps. To rule away this possibility, we recommend the authors study the variability in theta and lambda of conjunctive vs non-conjunctive grid cells. If the authors suspect that this phenomenon could explain part of their results, they should also take into account the findings of Gerlei and colleagues (2020) from the Nolan lab, that add complexity to this issue.

We appreciate the reviewer pointing out the possible role conjunctive cells may play. To investigate how conjunctive cells may affect the observed grid property variability, we have performed additional analyses taking into account if the grid cells included in the study are conjunctive. Comparing within- and between-cell variability of conjunctive vs. non-conjunctive cells in recording R12, we do not see any qualitative differences for either orientation or spacing (Fig S7A-B). When excluding conjunctive cells from the between-variability comparison, we do not see any significant difference compared to when these cells are included (Fig S7C-D). As such, it does not appear that conjunctive cells are the source of variability in the population.

We further note that the number of putative conjunctive cells varied across modules and recordings. For instance, in recording Q1 and Q2, Gardner et al. (2022) reported 3 (out of 97) and 1 (out of 66) conjunctive cells, respectively. Given that we see variability robustly across recordings (Fig 5), we do not believe that conjunctive cells can explain the presence of variability we observe.

(4) The results in Figure 6 are correct, but we are not convinced by the argument. The fact that grid cells fire in the same way in different parts of the environment and in different environments is what gives them their appeal as a platform for path integration since displacement can be calculated independently of the location of the animal. Losing this universal platform is, in our view, too much of a price to pay when the only gain is the possibility of decoding position from a single module (or non-adjacent modules) which, as the authors discuss, is probably never the case. Besides, similar disambiguation of positions within the environment would come for free by adding to the decoding algorithm spatial cells (non-hexagonal but spatially stable), which are ubiquitous across the entorhinal cortex. Thus, it seems to us that - at least along this line of argumentation - with variability the network is losing a lot but not gaining much.

We agree that losing the continuous attractor network (CAN) structure and the ability to path integrate would be a very large loss. However, we do not believe that the variability we observe necessarily destroys either the CAN or path integration. We argue this for two reasons. First, the data we analyzed [from Gardner et al. (2022)] is exactly the data set that was found to have toroidal topology and therefore viewed to be consistent with a major prediction of CANs. Thus, the amount of variability in grid properties does not rule out the underlying presence of a continuous attractor. Second, path integration may still be possible with grid cells that have variable properties. To illustrate this, we analyzed data from Sorscher et al. (2019) recurrent neural network model (RNN) that was trained explicitly on path integration, and found that the grid representations that emerged had variability in spacing and orientation (see point #6 below).

(5) In Figure 4 one axis has markedly lower variability. Is this always the same axis? Can the authors comment more on this finding?

We agree that in Fig 4 the first axis has lower variability. We believe that this is specific to the module R12 and does not reflect any differences in axis or bias in the methods used to compute the axis metrics. To test this, we have performed the same analyses for other modules, finding that other recordings do not exhibit the same bias. Results for the modules with the most cells are shown below (Author response image 1).

Grid propertied along Axis 1 are not less variable for many recorded grid modules.

Same as Fig.4C-D, but for four other recorded modules. Note that the variability along each axis is similar.

(6) The paper would gain in depth if maps coming out of different computational models could be analyzed in the same way.

We agree with the reviewer that examining computational models using the same approach would strengthen our results and we appreciate the suggestion. To address this, we have analyzed the results from a previous normative model for grid cells [Sorscher et al., (2019)] that trained a recurrent neural network (RNN) model to perform path integration and found that units developed grid cell like responses. These models have been found to exhibit signatures of toroidal attractor dynamics [Sorscher et al. (2023)] and exhibit a diversity of responses beyond pure grid cells, making them a good starting point for understanding whether models of MEC may contain uncharacterized variability in grid properties.

We find that RNN units in these normative models exhibit similar amounts of variability in grid spacing and orientation as observed in the real grid cell recordings (Fig S8A-D). This provides additional evidence that this variability may be expected from a normative framework, and that the variability does not destroy the ability to path integrate (which the RNN is explicitly trained to perform).

The RNN model offers possibilities to assess what might cause this variability. While we leave a detailed investigation of this to future work, we varied the weight decay regularization hyper-parameter. This value controls how sparse the weights in the hidden recurrent layer are. Large weight decay regularization strength encourages sparser connectivity, while small weight decay regularization strength allows for denser connectivity. We find that increasing this penalty (and enforcing sparser connectivity) decreases the variability of grid properties (Fig S8E-F). This suggests that the observed variability in the Gardner et al. (2022) data set could be due to the fact that grid cells are synaptically connected to other, non-grid cells in MEC.

(7) Similarly, it would be very interesting to expand the study with some other data to understand if between-cell delta_theta and delta_lambda are invariant across environments. In a related matter, is there a correlation between delta_theta (delta_lambda) for the first vs for the second half of the session? We expect there should be a significant correlation, it would be nice to show it.

We agree this would be interesting to examine. For this analysis, it is essential to have a large number of grid cells, and we are not aware of other published data sets with comparable cell numbers using different environments.

Using a sliding window analysis, we have characterized changes in variability with respect to the recording time (Figure S5A). To do so, we compute grid orientation and spacing over a time-window whose length is half of the total length of the recording. From the population distribution of orientation and spacing values, we compute the standard deviation as a measure of between-cell variability. We repeat the same procedure, sliding the window forward until the variability for the second half of the recording is computed.

We applied this approach to recording ID R12 (the same as in Figs 2-4) given that this recording session was significantly longer than the rest (almost two hours). Results are shown in Fig S5 B-C. For both orientation and spacing, no systematic changes of variability with respect to time were observed. Similar results were found for other modules (see caption of Fig S5 for statistics).

We also note that the rats were already familiarized with the environment for 10-20 sessions prior to the recordings, so there may not be further learning during the period of the grid cell recordings. No changes in variability can be seen in Rat R across days (e.g., in Fig 5B R12 and R22 have similar distributions of variability). However, we note that it may be possible that there are changes in grid properties at time-scales greater than the recordings.

Version 2

Abstract

Abstract

Although grid cells are one of the most well studied functional classes of neurons in the mammalian brain, whether there is a single orientation and spacing value per grid module has not been carefully tested. We analyze a recent large-scale recording of medial entorhinal cortex to characterize the presence and degree of heterogeneity of grid properties within individual modules. We find evidence for small, but robust, variability and hypothesize that this property of the grid code could enhance the encoding of local spatial information. Performing analysis on synthetic populations of grid cells, where we have complete control over the amount heterogeneity in grid properties, we demonstrate that grid property variability of a similar magnitude to the analyzed data leads to significantly decreased decoding error. This holds even when restricted to activity from a single module. Our results highlight how the heterogeneity of the neural response properties may benefit coding and opens new directions for theoretical and experimental analysis of grid cells.

Introduction

The discovery of grid cells in medial entorhinal cortex (MEC) [1] has led to considerable experimental and computational work aimed at identifying their origin [2–12] and their function [13–21]. The organization of grid cells into discrete modules [22, 23], with grid properties (grid spacing and orientation) clustered within module, but not between modules, has fundamentally shaped this research. For instance, the increasing size and spacing of grid modules along the dorsal-ventral axis of MEC, by discontinuous jumps of a near constant ratio, has been argued to be optimal for encoding local spatial information [24, 25] when grid cell activity across all modules is integrated together [24–28]. This is despite the fact that the individual neurons in hippocampus, a downstream target of the MEC, receive inputs from only a portion of the dorsal-ventral axis [29]. The modularity of the grid system has also been proposed to simplify the wiring necessary for generating continuous attractor dynamics [2, 3, 5, 16], a computational mechanism theorized to underlie grid cells function that enjoys considerable experimental support [23, 30–34].

Much of this prior theoretical work has made the additional assumption that grid cell properties are identical, up to a phase shift, within a single module [24, 25, 27, 28, 35]. However, as experimentalists have been aware, the distributions of measured orientation and spacing show non-zero variability [22, 30, 34]. This could be due to finite recording time, neurophysiological noise, and/or the sensitivity of numerical methods used to fit grid properties. Alternatively, this variability could reflect underlying inhomogeneity, at a fine scale, within modules. Despite the fundamental way in which grid cell function has been informed by their modular organization, to the best of our knowledge, no characterization of the degree and robustness of variability in grid properties within individual modules has been performed.

If robust variability of grid properties does exist, then it is possible that this heterogeneity could be exploited to encode additional information about local space, reducing the requirement of integration across multiple grid modules (Fig. 1). In particular, when all grid cells in a module have the same grid orientation and spacing, the joint population activity has a translational invariance that persists, even as larger populations of grid cells are considered (Fig. 1, “Fixed grid properties within module”). Thus, it is not possible to resolve spatial position from population activity. In contrast, if grid cells within a module have variability in their grid orientation and spacing, then – over a finite area – the translational invariance of the population activity is broken. In such a case, distinct patterns of population activity emerge in distinct locations of space (Fig. 1, “Variable grid properties within module”). As an example, the blue and green grid cells in Fig. 1B show the most overlap in the upper left half of the arena (denoted by cyan pixels), while the red and blue grid cells show the most overlap in the lower right portion of the arena (denoted by purple pixels). This is despite the fact that the variability in spacing and orientation is only on the order of a few centimeters and degrees, respectively. We expect that these patterns should become more complex and contain more information that could be used to disentangle spatial location as more grid cells are considered. In this paper, we perform detailed analysis of recent state-of-the-art MEC electrophysiological recordings, which were made publicly available [34]. We characterize the variability of grid orientation and spacing within individual modules and find evidence for small, but robust, variability. This variability is present whether a single value of grid orientation and spacing is assigned to each grid cell, or whether grid distortion [36, 37] is taken into account by considering three grid orientations and spacings independently. Performing similar analysis on recent normative recurrent neural network (RNN) models of grid cells [10], we find the presence of comparable variability. This provides another possible explanation for why these RNN models have been found to capture MEC grid cell response profiles [38]. To assess the functional implications of this heterogeneity, we perform simulation experiments with synthetic, noisy grid cell populations, where we have complete control over the distribution of grid orientation and spacing. We find that the variability in grid cell orientation and spacing, at a similar degree as present in the data we analyze, leads to lower decoding error of local space when using the activity of a single module.

Variability in grid cell properties within a module leads to enhanced encoding of local space.

When the activity of three idealized grid cells, all with the same grid spacing and orientation, are considered, the periodicity of the responses limits the amount of information conveyed about local space (Left column – “Fixed grid properties within module”). That is, there are multiple locations in physical space with identical population level activity. However, when three grid cells with variable grid spacing and orientation (in the realm of what is measured within individual grid modules – see Results), their joint activity contains considerably more information (Right column – “Variable grid properties within module”). This benefit of spatial inhomogeneity is expected to increase with larger populations of grid cells. Dashed squares in the joint activity map are enlarged below.

Taken together, our results challenge a frequently made assumption in the theoretical literature and support a growing understanding of the spatial information encoded by grid cell populations [39–42]. Additionally, our results encourage consideration of the broader benefits that multiple modules may provide, beyond the encoding of local space [43–45].

Results

Robust differences in grid cell properties within individual modules

To determine the extent to which variability of grid properties in individual modules exists, and to what extent this variability is a robust property of the grid code, we analyzed previously published MEC recordings [34], which include tens to hundreds of grid cells simultaneously recorded. This allows us to characterize the distribution of grid properties within a single grid module, to an extent not possible with other data sets. For each grid cell, we compute the grid spacing (λ) and orientation (θ) by measuring properties associated with the six hexagonally distributed peaks of the spatial autocorrelogram (SAC), as traditionally performed [46] (Fig. 2A; see Materials and Methods). For clarity, we begin by focusing on a single module, recorded from an open field environment (recording identifier: Rat R, Day 1, Module 2 – R12). This module was picked for its long recording time (approximately 130 minutes of recorded activity, as compared to the other open field recordings that have 30 minutes or less of recorded activity) and for its large number of simultaneously recorded putative grid cells (N = 168). In this module, we find that grid cells with high grid scores (> 0.85; N = 74) have a range of grid orientation and spacing (example cells shown in Fig. 2B; distributions across module shown in Fig. 2E, F), with λ ranging from 65 cm to 90 cm and θ ranging from 2° to 9°. Overlaying the SACs of pairs of grid cells with similar grid spacing and different grid orientation (Fig. 2C) or vice versa (Fig. 2D) enables visualization of the extent of this variability.

Grid properties are variable within a single grid module (recording ID R12).

(A) Overview of the standard procedure used to calculate the grid spacing and orientation of a given grid cell. First, spike maps are computed by identifying the location of the animal at the time of each spike. Gray line denotes the trajectory of the rat, red dots denote locations of spikes. A rate map is constructed by binning space and normalizing by the amount of time the rat spent in each spatial bin. A spatial autocorrelogram (SAC) is computed and, after the center peak is masked out (white pixels in the center of the spatial autocorrelogram – leading to change in color scale), the grid properties are fit by measuring the length and angle of the three peaks closest to 0°. (B) Example grid cells from the same module (recording ID R12), with estimated grid score, orientation (θ), and spacing (λ). (C)–(D) SAC overlaid for two pairs of grid cells [from (B)]; one pair with different θ and similar λ (C) and the other with similar θ and different λ (D). (E)–(F) Distribution of θ (E) and λ (F) across all grid cells with grid score > 0.85 (N = 74).

Because individual grid fields can be cut-off by the boundaries of the environment, it is possible that computing the grid spacing from the SAC (which considers all grid fields) could lead to an under-estimate of λ for some grid cells. To verify that the broad distribution of grid spacing, within the same module, that we see is not due to the specifics of the SAC, we recomputed the grid spacing for all grid cells directly from their rate maps, considering only the three grid fields closest to the center of the environment (Fig. S1A; see Materials and Methods). We find that while the grid spacing estimated from the SAC tends to be larger than the grid spacing estimated from the rate maps, a similarly broad range of λ is again present (Fig. S1B, C).

To assess whether the heterogeneity of grid properties present in a single grid module is a robust feature or attributable to noise (either in the recording or the grid property fitting procedure), we measure the variability in grid orientation and spacing within a single grid cell and between pairs of grid cells. If the heterogeneity is explainable by noise, then we expect that the within-cell variability will be of the same magnitude as the between-cell variability. In contrast, if the heterogeneity is a robust feature of the grid code, then we expect the within-cell variability will be significantly smaller than the between-cell variability. To measure the within- and between-cell variability, we split the recording into evenly spaced 30 second bins, randomly assigning each temporal bin to one of two equal length halves and computing the grid properties for each half of the recording (Fig. 3A; see Materials and Methods). We set inclusion criteria to filter out cells that do not have consistent hexagonally structured SACs across splits of the recording (see Materials and Methods). Although these requirements are strict (see Fig. S2 for percent of cells rejected), they set a conservative estimate on the amount of grid property variability, ensuring that we do not artificially inflate the variability due to inclusion of unreliable grid cells. We found that the length of the temporal bin used to split the data does not have a large impact on the percentage of cells accepted by this criteria (Fig. S3; see Materials and Methods).

Variability of grid properties is a robust feature of individual grid module (recording ID R12).

(A)Schematic overview of approach used to compute the between- and within-cell variability of grid orientation and spacing. (B)–(C) Distribution of within- and between-cell variability of θ and λ, respectively. Note that the distribution is across all 100 random shuffles of the data into two halves. (D) Average within-cell variability of grid orientation , compared to average between-cell variability of grid spacing (E) Same as (D), but for λ. 1 cell was excluded from (E) for visualization but was included in non-parametric statistical analysis. For (D)–(E), N = 82.

We find that the distribution, across 100 random shuffles of the data into two halves, of within-cell variability of grid orientation and spacing is more concentrated around 0 than the between-cell variability. Comparing the average within- and between-cell variability of grid spacing and orientation reveals that nearly all of the grid cells that passed the criteria for inclusion (N = 82) exhibit more between-cell than within-cell variability (Fig. 3D, E): 95.1% grid cells for orientation ; denotes mean across cells) and 100% of grid cells for spacing . A Wilcoxon-Signed-Rank Test indicates that between-cell variability is significantly greater than within-cell variability, for both orientation and spacing (Δθ: p < 0.001, Δλ : p < 0.001). We perform a number of control analyses to confirm that our results are robust, finding that: (1) the average grid field width does not significantly affect the between-cell variability (Fig. S4), (2) the amount of between-cell variability does not change significantly across the recording duration (Fig. S5), (3) the between-cell variability is not significantly impacted by the boundaries of the arena (Fig. S6), and (4) the amount of between-cell variability is not driven by the presence of cells that have both head direction and grid cell like tuning (“conjunctive” cells [46] – Fig. S7).

In keeping with convention, the reported grid cell properties are the average of those computed for each of the three independent axes in the SAC (Axis 1: aligned to ≈ 0°; Axis 2: aligned to ≈60°; Axis 3: aligned to ≈ ™ 60° [47]). To ensure that this averaging is not contributing to greater between-cell variability, we repeated the analysis above, restricting ourselves to each axis separately. The results again demonstrate that grid properties are significantly more robust within-cell than between-cell (Fig. 4A, B). For each axis, the average between-cell variability for every cell was significantly higher than the average within-cell variability (Fig. 4C, D), as reported by the Wilcoxon-Signed-Rank Test for orientation (p < 0.001, for all three axes) and for spacing (p < 0.001, for all three axes).

Variability of grid properties, restricted to the same axis, is a robust feature of individual grid module (recording ID R12).

Same analysis as in Fig. 3 (B) – (E), but for variability measured on each axis independently. For visualization, we exclude a small number of cells that were outside the axes limits, including 2, 5, and 10 cells for Axes 1–3, respectively (C); and 3, 4, and 3 cells for Axes 1–3, respectively (D); these cells were included in non-parametric statistical analyses.

Having demonstrated that grid cell properties are robustly heterogeneous in a single module, we proceed to analyze the remaining recordings in the data set (N = 420 cells across 8 modules; one module in the data set had no cells that passed our criteria) [34]. Although there are differences across recordings, we find larger between-than within-cell variability for grid orientation and spacing is present across all recordings (Fig. 5A, B; 80.0% of grid cells have greater between- than within-cell variability for orientation, of grid cells have greater between- than within-variability for spacing, and ). A Wilcoxon-Signed-Rank Test finds that these differences are significant (Δθ: p < 0.001; Δλ : p < 0.001). We do not find evidence suggesting that the within-cell variability is influenced by grid score (Fig. 5C, D; linear regression for θ: R2 = 0.03, p = 0.55 Wald Test; linear regression for λ: R2 = 0.07, p = 0.14 Wald Test).

Within module grid property variability is a robust feature across modules.

(A) Average within-cell variability of grid orientation , compared to average between-cell variability of grid orientation for each cell (N = 420) across 8 modules (cells colored by their corresponding recording ID). The histogram above the plot shows the distribution of and the histogram to the right shows the distribution of (B)Same as (A), but for grid spacing. For visualization, 5 cells are excluded , but are included in non-parametric statistical analyses. Dashed gray lines show the population mean. (C)–(D) Average within cell variability of θ and λ (respectively), as a function of grid score. For visualization, 3 and 22 cells are excluded from (C)–(D), respectively, but are included in statistical analyses. Black solid line is linear regression, with R2 and p-value reported above.

To understand whether the observed grid cell variability may be a heretofore unknown property of MEC computational models, we trained recurrent neural networks (RNN) to perform path integration. These RNN models have previously been shown to develop grid responses [7, 8, 10] and have been argued to develop continuous attractor network structure [12], making them normative models for MEC function. We find that the RNNs develop grid responses with a distribution of grid spacing and orientation (Fig. S8A–D), much like the neural data we analyzed (Fig. 2). Changing the sparsity of the recurrent layer by increasing the strength of weight decay regularization reduces the variability (Fig. S8E–F).

Taken together, our analysis of large-scale MEC recordings demonstrates that grid cells in the same grid module do not have a single grid spacing and orientation, but instead have a restricted range of values around the module mean. This property is a robust feature that cannot be explained by noise from the recording or fitting procedures and emerges in existing normative models of MEC.

Variability in grid properties within individual modules improves the encoding of local space

The variability of grid cell properties within individual grid modules, while statistically significant, is small in magnitude. Can a computational benefit in the encoding of local space be gained from this level of inhomogeneity? How sensitive might such a computational benefit be to the exact amount of variability present?

To address these questions, we generate populations of synthetic grid cells (see Materials and Methods), where we have complete control over the number of grid cells and their firing properties. For simplicity, we assume that all grid cells in our population have grid orientation and spacing sampled from Gaussian distributions, with means and and standard deviations σλ and σθ, respectively. Assigning each grid cell in our population a grid orientation and spacing, we are able to generate “ideal” rate maps [14]. Sampling from a Poisson process on these ideal rate maps, we generate noisy grid cell rate maps (Fig. 6A). Using a simple linear decoder (see Materials and Methods), we can examine how decoding error of local space is affected by the number of grid cells in the population and the amount of variability (σλ and σθ).

Variability in grid properties enables improved decoding of local space from the activity of grid cells within a single module.

(A) Example noisy grid cell rate maps generated from a Poisson process. The size of the square arena is set to 1.5 m × 1.5 m to be consistent with what was used in the experimental set-up analyzed [34]. (B)–(C) Distribution of sampled grid spacing and orientation from synthetic population, when using , , σλ = 5 cm, and σθ = 1°; compare to the distribution measured from real data (Fig. 2E, F). (D) Decoding error, as a function of grid cell population size, with populations having either no variability in grid properties (black line) or variability similar to what was present in the data analyzed (blue line). The solid line is the mean across 25 independent grid cell populations and the shaded area is ± standard deviation of the 25 independent populations. The dashed black line shows chance level decoding error. (E) Decoding error for synthetic populations and real data for up to N = 64 cells (red line). (F) Decoding error, over a grid of σθ and σλ values, for populations of N = 1024 grid cells. White star denotes values used in (D).

We begin by investigating the decoding capabilities of a synthetic module with properties similar to that of the experimentally recorded module that was analyzed in detail (recording identifier R12; Figs. 2, 3). We therefore set and . To determine an appropriate value for σλ and σθ, we subtract the mean between-cell variability by the mean within-cell variability and . We interpret these values as the amount of variability in grid spacing and orientation that is not due to noise. Therefore, we set σθ = 1° and σλ = 5 cm. This amount of variability leads to similar sampled distributions of λ and θ as was found in the real data (compare Fig. 6B, C with Fig. 2E, F).

Applying the linear decoder to populations of synthetic grid cells, we find that decoding error decreases as the number of grid cells is increased (Fig. 6D, blue line). When N = 1024, the decoding error is ≈ 20 cm, substantially better than random decoding (Fig. 6D, dashed black line). As expected, this result is not seen in a synthetic population of grid cells with zero variability in grid properties (σθ = 0° and σλ = 0 cm) (Fig. 6D, black line). In particular, while the restricted number of grid firing fields enables a decoding error smaller than chance, the population with fixed grid properties exhibits little change in decoding error with increasing numbers of grid cells. This demonstrates that the improved encoding is specific to populations with inhomogeneity in their grid spacing and orientation.

To validate that the synthetic population is a reasonable surrogate to the experimentally recorded data, we perform the same decoding analysis on grid cells from recording ID R12 (see Materials and Methods). As the real data is limited in the number of cells, we are only able to compare up to populations of N = 64 (Fig. 6E). However, in this restricted range, we find agreement between the synthetic population with grid property variability and the recorded data (Fig. 6E, compare red and blue lines). This supports our hypothesis that the observed amounts of inhomogeneity in grid spacing and orientation can lead to improved decoding of local space.

To determine the extent to which the decrease in decoding error depends the exact variability of grid spacing and orientation, we perform a grid search over 36 pairs of (σλ, σθ) values in [0°, 1°, …, 5°] × [0 cm, 1 cm, …, 5 cm]. As expected, when the variability of both grid properties is large, we find nearly 0 cm decoding error of local space (Fig. 6E, bottom right). However, there is additionally a range of σλ and σθ values that lead to improved decoding, including values smaller than the values matched to the experimental data (Fig. 6E, above and to the left of the white star). We perform the same grid search on synthetic populations of grid cells with different spacings, and (Fig. S9A, B), finding again that the decoding error drops below that of the fixed population with sufficient grid property variability. For the synthetic module with smaller spacing , the existence of more grid fields in the 1.5 m × 1.5 m arena leads to greater complexity of interference patterns (Fig. 1), enabling the same amount of variability in grid spacing and orientation to lead to a sharper decrease in decoding error. Additionally, for the synthetic module with smaller spacing, a given value of σλ is a larger percentage of , as compared to the synthetic module with larger spacing.

Finally, we consider how decoding within a single module compares to decoding across two modules. When the two modules are consecutive (e.g., modules 1 and 2), their mean grid spacing, and , has been experimentally found to be related via [22]. In such a case, decoding activity from populations with fixed grid properties leads to nearly 0 cm error (Fig. S10A, black line). The addition of variability in grid properties does not significantly change the behavior of the decoding (Fig. S10A, blue line). This suggests that small amounts of inhomogeneity may not disrupt the previously achieved theoretical bounds on decoding from multiple grid modules [24, 25, 28, 48]. However, if the two modules being decoded from are non-consecutive (e.g., modules 1 and 3) then the mean grid spacing can be related by an integer multiple, . In such a setting, the grid fields of the larger module are a subset of the grid fields of the smaller module, up to a rotation (due to the difference in orientation between the two modules), and we again find that variability in grid properties can improve decoding accuracy (Fig. S10B, compare black and blue lines). Similar results are found when using experimentally found values of [22] and not assuming an exact relationship.

Taken together, these results suggest that individual grid modules can exhibit significant encoding of local space via heterogeneity in their grid properties, even when the extent of the variability in θ and λ is similar to that found in the analysis of the experimental recordings. This benefit can also improve encoding in cases when multiple, non-consecutive modules are considered.

Discussion

The multiple firing fields of grid cells, organized along a triangular lattice, has been historically interpreted as a limiting feature for the encoding of local space. Particularly influential in shaping this view has been the discovery of the distribution of grid cells into distinct modules, with grid cell spacing (λ) and orientation (θ) preserved within, but not across, modules [22, 23], making integration across multiple modules necessary for spatial information to be decoded [24–28]. While evidence for discontinuity in the grid cell properties across modules is strong, the corollary assumption, that within-module values of λ and θ are identical (up to the bounds of noise), has not been systematically studied.

Analyzing recently collected MEC recordings, we found the range of λ and θ values was large, with examples of grid cell pairs in the same module having over 5° difference in grid orientation and 20 cm difference in grid spacing (Fig. 2). Statistical analysis shows that the variability is more robust than expected from noise, for the majority of grid cells (Fig. 5). This was despite the fact that we used a very conservative criteria for assessing whether a grid cell was consistent enough to be included in the analysis. Our comparison of within- and between-cell grid property variability was key to our argument, as it was for previous work using it to study the robustness of differences in peak grid field firing rates [41]. The absence of its use in the characterization of distribution of λ and θ [30] may be why the consistency of this heterogeneity was not identified until now. We find that our conclusion holds when performing a number of control experiments (Figs. S1, S4–S7) and whether we treat each grid field independently or take the average across grid fields (Figs. 3, 4). This challenges the assumption that the variability observed in the grid orientation and spacing is attributable solely to measurement noise.

We find that normative recurrent neural network models that develop grid cells when optimized to perform path integration [7, 8, 10] develop similar amounts of grid property variability (Fig. S8). This illustrates the ability of these RNN models to capture aspects of grid cell properties that have not been previously studied, and may provide another explanation for why these models have greater similarity to real MEC recordings than other models [38]. Probing these computational models in greater depth may enable a more detailed understanding of the observed grid property heterogeneity.

We hypothesized that this variability may be used to increase the fidelity at which individual grid modules can encode local space. This idea is consistent with a large body of literature showing that heterogeneity in the responses of populations of neurons increases the robustness of encoding [49–52]. We find, in noisy synthetic populations of grid cells, that a level of variability in grid properties similar to what is quantified in the real data can be sufficient to accurately decode information of local space (Fig. 6). This benefit is increased with larger numbers of grid cells in the population (Fig. 6D) and is observed over a range of values for the underlying variability (Fig. 6F). We find that the improvement was most pronounced in modules with small grid spacing (Fig. S9A), although larger modules can see a decrease in decoding error for amounts of variability consistent with was was found in the analyzed data (Fig. S9B).

We note that our results are additionally aligned with recent findings of heterogeneity in maximal firing rates across individual grid fields [39–41] and grid sheering [36, 37, 42]. Our work further demonstrates that, even in the absence of these perturbations, individual grid modules may encode considerably more local spatial information than previously believed.

Finally, models of the formation of orientation maps in visual cortex have demonstrated that slight angular offsets of retinal mosaics, along which retinal receptive fields are organized, can generate complex patterns [53] similar to those found in visual cortex orientation maps [54]. Our results indicate that grid cells in MEC may take advantage of a similar computational principle, suggesting that mosaic patterns might be a broadly utilized feature of neural coding [55].

Limitations

While the data set we analyzed [34] represents an advance in the ability to simultaneously record from tens to hundreds of putative grid cells, across grid modules, the MEC remains a challenging brain region to access for large-scale neurophysiological experiments. Indeed, with our conservative inclusion criteria, we were ultimately limited by having only 420 grid cells included in our analysis. Future work can perform more detailed and complete characterizations of grid property heterogeneity, as new neurotechnologies that enable larger yield of grid cells are developed [56, 57].

Our decoding analysis, while demonstrating the possibility that variability in grid properties can be used by individual grid modules to enhance the encoding of local spatial information, made several simplifying assumptions, including: (1) independent Poisson noise for neural activity, (2) linear decoding, and (3) normal distribution of grid properties. Although comparison to real data showed that these assumptions are reasonable (Fig. 6E), future work can assess the extent to which these restrictions can be lifted (e.g., to incorporate correlated neural noise), while still enabling individual grid modules to have low decoding error.

Open questions

The heterogeneity in grid properties we characterize motivates the investigation of several new lines of research. Because these are directions that we believe to be fruitful for the field as a whole, we outline them below, with our hypotheses for possible answers.

Q1: How does grid property variability affect continuous attractor network structure?

Continuous attractor network models of grid cells [2, 3, 5, 16] enjoy considerable experimental support [23, 30–34], making them one of the “canonical” models in neuroscience. However, these models make use of the assumption that all the grid cells in a given module have the same grid orientation and spacing to simplify the network connectivity. In particular, by assuming equal grid orientation and spacing, it becomes possible to arrange grid cells in a two-dimensional space spanned by their phases (i.e., a “neural sheet” [16]). Neurons close in this space are wired with excitatory connections and neurons far in this space are wired with inhibitory connections. As the data set we analyzed was found by others to provide strong support for the basic predictions of continuous attractor networks (i.e., toroidal topology of the activity manifold) [34], we do not view our results as directly challenging these models. That the RNN models we investigate (Fig. S8) are explicitly trained to perform path integration (and achieve highly accurate performance [12]) supports our hypothesis that variability in grid properties does not necessarily destroy the continuous attractor or path integration capabilities of the MEC. Understanding how this is possible is an exciting future direction, and use of geometric [58] and dynamical systems based tools [59–61] may shed new light on this. We hypothesize that the degree in variability of grid spacing and orientation may strike a balance between being small enough to keep the continuous attractor network structure stable, but large enough to enable encoding of local information of space.

Q2: What causes grid property variability?

A natural direction to address is identifying the source of the heterogeneity in grid cell properties we observe. One hypothesis is that this could be driven by “defects” in the specific connectivity pattern that is needed for a continuous attractor. This could be due to synaptic connections between grid cells and other functional classes of neurons in the MEC, such as border [62], band [63], and non-spatial [39] cells. Because the path integrating RNN models have been found to develop analogous responses to these classes [7, 8, 64–66], as well as capture general properties of MEC activity [38], we examined the affect of “connectivity noise” in RNNs. In particular, we swept across different strengths of weight decay regularization, which controls the amount of sparsity in the recurrent layer. We find that smaller weight decay (and thus, denser connectivity between units) led to greater variability in grid properties (Fig. S8E, F). Increasing the weight decay (and thus, enforcing sparser connectivity) led to a reduction of grid property variability (Fig. S8E, F). This supports the hypothesis that the heterogeneity in grid properties we observe may be due to “non-perfect” connectivity.

Alternatively, the coupling between hippocampus and MEC, which has been shown to lead to variability in grid field firing rates (as experimentally observed) [41, 67], may lead to differences in grid orientation and spacing. In particular, a previous computational model that learned grid cells from place cell input, using non-negative principal component analysis, has shown that place field width affects grid cell orientation and spacing [6]. Heterogeneity in the spatial coding properties of place cells has been found along the transverse axis of CA3 [68–70], suggesting there may be a systematic differences in the place field widths of hippocampal inputs to MEC grid cells. Further, it was shown that this place-to-grid cell computational model has a linear relationship between place field width and grid spacing, and a non-monotonic relationship between place field width and grid orientation [6]. These may explain why we find stronger average variability in grid spacing than grid orientation (Fig. 5A, B).

Q3: How does grid property variability shape hippocampal representations?

The projections from MEC to hippocampus suggest that the variability in grid properties may influence hippocampal representations (even if grid cells do not comprise the majority of its inputs [39]). We consider two possible ways in which this may happen. First, given that grid cells have been reported to maintain their grid spacing and orientation across exposures to new environments, while undergoing a change in their grid phase [30, 71], the integration across multiple modules has been necessary to explain place field remapping. However, grid phase plays an important role in generating the specific complex interference patterns that emerge when considering the joint activity of grid cells with variable grid properties (Fig. 1). Thus the reported changes in phase may be sufficient to generate large (and seemingly random) changes in local spatial information conveyed by grid cells to hippocampal cells. This could drive additional changes in the local spatial information projected to hippocampus, as well as explain the significant differences in correlation structure between CA1 neurons across different environments [72].

And second, recent work on hippocampal place field drift [73–77] has demonstrated that there is a significant change in the place field location across time, especially in CA1. One possible source of this phenomenon is the reported instability of dendritic spines on the apical dendrites of CA1 place cells [70, 78– 80], ostensibly leading to changes in the MEC inputs to these neurons. However, if grid cells across multiple modules are necessary for local spatial information, turnover in synaptic input is unlikely to cause large changes in the spatial preferences of CA1 neurons, as integration over several scales should provide stable encoding properties. In contrast, different subpopulations of grid cells with variable grid properties can lead to differences in the local spatial information encoded by their joint activity, even if they come from a single module, possibly influencing the spatial preferences of CA1 place cells.

Q4: Why are there multiple modules?

Given that our results demonstrate the ability of single grid modules to encode information about local space – a feat previously believed to be possible only if activity from multiple grid modules was integrated together – why is does MEC have multiple modules? While the variability removes the necessity for encoding local space with multiple modules, higher fidelity representations is achievable by integrating across multiple modules (Fig. S10). In addition, the use of grid cells beyond spatial navigation [81–87], where hierarchical representations are important [44, 45], may be a sufficient implicit bias for the formation of multiple modules [88]. In particular, encoding information at multiple distinct scales is critical for multi-scale reasoning, a cognitive function grid cells may support.

Materials and Methods

Electrophysiology recordings

The neural activity analyzed in this paper comes from a publicly available data set, which has previously been described in detail [34]. We provide brief summary of the methodology and the experimental paradigms used during the recordings.

Three male rats (Long Evans – Rats Q, R, and S) were implanted with Neuropixels silicon probes [89, 90]. These probes were targeted at the MEC-parasubiculum region and the surgery was performed as described previously [32, 90]. After three hours of recovery, recordings were performed.

In the recordings analyzed, the rats foraged for randomly dispersed corn puffs in a 1.5 × 1.5 m2 square open field arena, with walls of height 50 cm. The rats were familiar with the environment and task, having trained 10–20 times prior to the implantation. The rats were food restricted to motivate their foraging, being kept at a minimum of 90% of their original body weight (300–500 grams).

All procedures in the original study were approved by the Norwegian Food and Safety Authority and done in accordance with the Norwegian Animal Welfare Act and the European Convention for the Protection of Vertebrate Animals used for Experimental and Other Scientific Purposes.

Electrophysiology post-processing

The neural activity analyzed in this paper was post-processed, before made publicly available. We describe, in brief, the post-processing performed [34], as well as the post-processing we performed on the downloaded data.

Spike sorting, via KiloSort 2.5 [90], was applied to the data recorded from the Neuropixel probes. Individual units were deemed putative cells if their average spike rate was in the range of 0.5–10Hz, and 99% of their interspike intervals were greater than 2 ms.

For each putative cell, rate maps were constructed by averaging the activity at binned spatial positions in the open field arena. This raw rate map was smoothed, using a Gaussian kernel. The autocorrelation of these rate maps were computed, and a grid score calculated, as described previously [46].

From the downloaded spike trains, we constructed rate maps and autocorrelograms in a similar manner, using code made publicly available [8] (Fig. 2).

Grid module classification

The public data set we analyzed [34] contained the module identity of all putative grid cells. In brief, these module identities were assigned by first projecting the 2D autocorrelogram of every recorded unit onto a 2D space using the non-linear dimensionality reduction algorithm UMAP [91]. Then, the DBSCAN algorithm [92] was used to cluster the units, based on their position in the dimensionally reduced space. Cluster membership served as the basis for grid module classification, with the largest cluster being removed as it was found to not contain spatially selective autocorrelograms. This non-supervised module assignment yielded clusters of high grid scores and similar grid spacing and orientation within each cluster-module. More details can be found in Gardner et al. (2022) [34]. We performed no further analysis regarding module identity.

Computing grid score, orientation and spacing from the spatial autocorrelogram

Grid spacing and grid orientation were computed according to standard methods described in detail previously [22, 93]. Briefly, the goal of the procedure is to identify the location of the six nearest fields in the spatial autocorrelogram (SAC) (Fig. 2A). This was achieved by performing the following steps. First, the SAC was smoothed using a 2D Gaussian filter with σ = 1. Second, the center of the SAC was excluded by applying a mask. Because the size of the SAC’s central peak changes for every module, the radius of such mask was re-computed for each module. Third, we thresholded the SAC by applying an extended-maxima transform ndimage.maximum_filter. Fourth, we identified the center of every field by using the function scipy.stats.find_peaks.

Once the peaks of every field had been found, we computed the location of every peak in polar coordinates. We then selected the 6 peaks that were closest to the center of the SAC, based on the computed radial components. Because every SAC is symmetric, we considered for further analysis the 3 peaks closest to the X axis in angular distance (Axis 1, 2, and 3 [47]). Grid spacing was computed as the arithmetic mean of the radial component of the 3 peaks (except when each peak was analyzed separately – Fig. 4). Given that the SAC dimensions are twice of that of the real arena, we multiplied the SAC radial mean by a factor of 2. Grid orientation was computed as the angular mean of orientations (relative to the x-axis) of the three peaks (except when each peak was analyzed separately – Fig. 4).

In order to ensure that subsequent analysis was performed only on cells whose SAC’s could be well described by hexagonal structure, we imposed the following constraints: (1) the relative angle between two peaks could not be < 30°; (2) the relative angle between two peaks could not be > 90°; (3) the values for grid spacing between the peaks could not be substantially different (the ratio of spacings between any two peaks must be > 0.5 and < 2). Cells that did not meet these criteria were determined to have “Poor grid fit” and were rejected from all subsequent analysis. The percentage of all cells that were removed by this inclusion criteria is shown in Fig. S2. We note that cells from one of the nine modules in the publicly available data set that we analyzed had 98.4% of cells rejected by this criteria (recording identifier – R23). We therefore did not include it in any subsequent analysis.

To compute the grid score of recorded MEC cells, we made use of previously published code [8], that is based on metrics that have become standards in quantifying grid cell properties[46]. For analysis of the distribution of grid properties (Fig. 2E, F), we included only grid cells with grid scores greater than 0.85. This was done to demonstrate that variability was present even in cells that exhibit robust grid cell properties. In the subsequent analyses, an alternative criteria is used, which considers the reliability of the grid responses (see below).

Computing spacing from the rate maps

In order to characterize the spacing with a method that is less susceptible to the effects of having grid fields at the boundaries of the environment (as the SAC method could be), we compute the spacing using the rate maps of individual grid cells (Fig. S1A). Once the rate map was computed, we smoothed it with a Gaussian filter, setting σ = 1.5. Then, using the function scikit-image.feature.peak_local_max we extract the position of the center of each firing field. Because our goal with this analysis is to show that the fields in the border do not impact our analysis with the SAC, we restrict ourselves to the 3 peaks that are closest to the center of the arena (Fig. S1A). We compute the distances between those three peaks to each other and report the spacing as the average of the distances measured.

Within and between cells splits

To characterize the within- and between-cell variability of grid spacing and orientation, we employed the following approach. First, we split the data into bins of fixed length (30 seconds). From this, we randomly assigned each interval to one of two blocks (denoted as blocks A and B), with exactly half the total number of intervals in each block. For each grid cell, we computed the grid spacing [ and ] and orientation [ and ], from the data in each block. The within-cell differences of grid spacing and orientation was determined as and. Each grid cell’s properties were also compared those of another grid cell, with the match being made using random sampling without replacement. These comparisons were determined as and . This process was repeated 100 times (referring to each iteration as a “shuffle”), per recording. Examples of splits of the data, for different shuffles, is schematically illustrated in Fig. 3A). The resulting distributions of Δθwithin, Δλwithin, Δθbetween, and Δλbetween, across grid cells and splits of the data were then compared. If the within-cell variability in grid spacing and orientation was smaller than between-cell variability, we concluded that the variability of grid cell properties was a robust feature of the data and not due to noise.

When performing this shuffle analysis, we found that some cells, despite having a good grid fit when all the data was considered, did not have SACs that were well described by hexagonal structure when the data was split in half. We viewed this a manifestation of unreliable grid coding and a possible confound in our quantification of variability. As such, we introduced a new inclusion criteria (replacing that of requiring a grid score of > 0.85), only considering cells that had poor grid fits on < 5% of all shuffles. The percentage of all cells that were removed by this inclusion criteria is shown in Fig. S2. In general, we found that cells with high grid score were reliable, although there were exceptions. Additionally, we found that size of the bin used for splitting the data did not significantly affect the percent of cells with good grid fits (passed the prior inclusion criteria) that were considered reliable (Fig. S3).

Synthetic grid cells

To study how variability in grid cell properties might endow the grid code with computational advantages, we generated synthetic grid cell rate maps, so that we could have complete control over the distribution of their properties. These synthetic grid cell rate maps were constructed as follows.

First, the lengths of each dimension the “arena” within which the simulated grid cells exist (Lx and Ly) were set. Then, for N ∈ ℕ grid cells, the grid spacing and orientation were sampled via λ(i) ∼ 𝒩 (µλ, σλ) and θ(i) ∼ 𝒩 (µθ, σθ) where 𝒩 (µ, σ) is a normal distribution with mean µ and variance σ2. Grid phase was sampled as ϕ(i) ∼ 𝒰 ([0, Lx] [0, Ly]), where ϕ is a two dimensional vector, with first component uniformly sampled from [0, Lx] and second component uniformly sampled from [0, Ly]. To construct a population with no variability in grid properties (to use as a control), we set σλ = 0 m. and σθ = 0°.

For each grid cell, we generated idealized grid responses by summing three two-dimensional sinusoids [14], such that the activity at x = (x, y) ∈ [−Lx/2, Lx/2] ×[−Ly/2, Ly/2] is given by where is the maximal firing rate, R(θi) is the two-dimensional rotation matrix, with rotation θi, and k j are the wave vectors with 0°, 0 60 °, and angular 120° differences [i.e., k1 = (1, 0)T,]. To match the recorded neural data, where individual grid cells have distinct maximal firing rates,. We enforced to be within [2, 30], by setting any sampled values outside of this range to the boundary values (i.e., 2 or 30).

To determine the extent to which local spatial information can be decoded from the activity of populations of grid cells with different degrees of variability in their grid properties, we performed the following analysis.

We generated noisy synthetic spike rates of N grid cells by assuming a Poisson process and sampling using the idealized rate maps (Eq. 1). More concretely, the activity of grid cell i at position (x, y) was assumed to be a random variable with a Poisson distribution, whose mean was X(i)(x, y). Thus, the probability of observing spikes, at position (x, y), is given by

Linear decoding synthetic data

For a given resolution of the arena, we generated , where j = 1, …, 10. That is, we constructed 10 noisy rate maps. We performed cross-validated decoding by averaging across 9 of the 10 rate maps, to get an average rate map . For sake of simplicity, consider . To decode local position from the held out noisy rate map [e.g. ], we multiplied the activity at each position by all the positions in the average rate map, taking the sum and assigning the decoded position as that with the largest value. The Euclidean distance between the decoded position and the true position is considered the error. This was performed 10 times (holding out each rate map once) and the average error across all positions in the environment was then averaged across all 10 of the validation splits.

Linear decoding experimental data

To decode the electrophysiological data [34], we sampled subpopulations of grid cells from the N = 82 units (recording ID R12) that passed the criteria described previously. For each subpopulation, we split the recorded activity into 10 evenly spaced temporal bins, and constructed average ratemaps for each bin. This is consistent with what was done in decoding the synthetic data. All ratemaps, for each cell, were normalized by the maximal activity. We then randomly chose 5 of the 10 time bins to serve as the “train” data, the remaining 5 served as the “test” data. The rate maps were averaged and then the linear decoder was applied. We sampled 10 different choices of splitting the data and 25 choices of subpopulation.

Recurrent neural network model

Code used for training and evaluating path integrating recurrent neural network (RNNs) [10, 12] was pulled from https://github.com/ganguli-lab/grid-pattern-formation. We used the same hyperparameters as are present on the repository (see Table 1), except when examining the effect of weight regularization on grid property variability, in which case we trained RNNs with weight decay magnitude from {10−6, 10−5, 10−4, 2.5 ·10−4 }. The greater the value, the greater the enforced sparsity. These were chosen as they had previously been found to lead to good path integration performance and have high grid scores [12]. We independently verified that this was true for the resulting RNNs. We computed the grid score, spacing, and orientation using similar code implementations as was used for the neural data. For each value of weight decay, we trained three independent RNNs, using different random seeds.